---
title: "Treatment Efficacy Analysis of a Randomised Trial"
subtitle: "Does rectal indomethacin prevent post-ERCP pancreatitis? Effect size, NNT and subgroups"
author: "Kingsley Amegah"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5.2, fig.align = "center")
suppressMessages({
  library(dplyr); library(forcats); library(broom)
  library(gtsummary); library(ggplot2); library(scales)
})
theme_set(theme_minimal(base_size = 12))
PAL <- c("Placebo" = "#9ec7bd", "Indomethacin" = "#2a9d8f")
ACC <- "#2a9d8f"; CORAL <- "#e76f51"; INK <- "#1d3a34"
# gtsummary fills empty cells with an em dash that, under the render's C locale,
# prints as the ASCII text "<U+2014>". We strip only that literal text pattern
# (pure ASCII, so it can never touch the data digits) to keep tables dash-free.
strip_dash <- function(x) gsub("&lt;U\\+[0-9A-F]{4}&gt;|<U\\+[0-9A-F]{4}>", "", x)

dat <- read.csv("data/indo_clean.csv", stringsAsFactors = TRUE)
dat$rx <- factor(dat$rx, levels = c("Placebo", "Indomethacin"))

# Effect measures from a 2x2 (treated vs control), with Wald 95% CIs.
eff <- function(a, b, c, d) {
  # a = events treated, b = non-events treated, c = events control, d = non-events control
  n1 <- a + b; n0 <- c + d; p1 <- a / n1; p0 <- c / n0
  rr <- p1 / p0
  se_lrr <- sqrt(1/a - 1/n1 + 1/c - 1/n0)
  or <- (a / b) / (c / d)
  se_lor <- sqrt(1/a + 1/b + 1/c + 1/d)
  arr <- p0 - p1
  se_arr <- sqrt(p1 * (1 - p1) / n1 + p0 * (1 - p0) / n0)
  list(p1 = p1, p0 = p0,
       rr = rr, rr_l = exp(log(rr) - 1.96 * se_lrr), rr_u = exp(log(rr) + 1.96 * se_lrr),
       or = or, or_l = exp(log(or) - 1.96 * se_lor), or_u = exp(log(or) + 1.96 * se_lor),
       arr = arr, arr_l = arr - 1.96 * se_arr, arr_u = arr + 1.96 * se_arr,
       nnt = 1 / arr, nnt_l = 1 / (arr + 1.96 * se_arr), nnt_u = 1 / (arr - 1.96 * se_arr))
}
```

# Background and question

A randomised controlled trial can do more than show that a treatment works. It
can quantify **how much** it helps, in terms a clinician and a patient can act
on: the absolute risk reduction, the relative risk, and the **number needed to
treat (NNT)**, the number of patients you must treat to prevent one event.

This report analyses a real, published placebo-controlled trial (Elmunzer and
colleagues, *New England Journal of Medicine*, 2012). **602 patients** at high
risk of complications from ERCP (a procedure on the bile and pancreatic ducts)
were randomised to a single dose of **rectal indomethacin** or **placebo**. The
primary endpoint was **post-ERCP pancreatitis (PEP)**, a painful and sometimes
serious inflammation of the pancreas.

# Baseline balance

Randomisation should make the two groups comparable before treatment. We check
this first.

```{r baseline, results='asis'}
bt <- dat |>
  select(rx, age, gender, risk, risk_group, prior_pep, sod, psphinc, difcan, train) |>
  tbl_summary(
    by    = rx,
    type  = list(gender ~ "dichotomous", risk_group ~ "dichotomous"),
    value = list(gender ~ "Male", risk_group ~ "Higher risk (>2.5)"),
    label = list(age ~ "Age, years", gender ~ "Male sex",
                 risk ~ "Risk score", risk_group ~ "Higher risk (score >2.5)",
                 prior_pep ~ "Prior pancreatitis", sod ~ "Sphincter of Oddi dysfunction",
                 psphinc ~ "Pancreatic sphincterotomy", difcan ~ "Difficult cannulation",
                 train ~ "Trainee involved"),
    digits = all_categorical() ~ c(0, 1)) |>
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 3)) |>
  modify_header(label ~ "**Characteristic**") |>
  modify_caption("**Table 1. Baseline characteristics by treatment arm**")
cat(strip_dash(as_kable(bt)), sep = "\n")
```

The arms were well matched on age, sex, risk score and every procedural risk
factor (all p-values large), confirming that randomisation produced comparable
groups. Any difference in pancreatitis can therefore be attributed to the
treatment.

# Primary efficacy: how big is the benefit?

```{r primary}
tab <- table(dat$rx, dat$pep)              # rows: Placebo, Indomethacin
a <- tab["Indomethacin", "Yes"]; b <- tab["Indomethacin", "No"]
c <- tab["Placebo", "Yes"];      d <- tab["Placebo", "No"]
e <- eff(a, b, c, d)
chip <- chisq.test(tab)$p.value
res <- tibble::tibble(
  Measure = c("PEP, placebo", "PEP, indomethacin",
              "Absolute risk reduction (ARR)", "Relative risk (RR)",
              "Relative risk reduction", "Odds ratio (OR)",
              "Number needed to treat (NNT)"),
  Value = c(sprintf("%.1f%%", e$p0 * 100),
            sprintf("%.1f%%", e$p1 * 100),
            sprintf("%.1f pp (95%% CI %.1f to %.1f)", e$arr*100, e$arr_l*100, e$arr_u*100),
            sprintf("%.2f (95%% CI %.2f to %.2f)", e$rr, e$rr_l, e$rr_u),
            sprintf("%.0f%%", (1 - e$rr) * 100),
            sprintf("%.2f (95%% CI %.2f to %.2f)", e$or, e$or_l, e$or_u),
            sprintf("%.0f (95%% CI %.0f to %.0f)", e$nnt, e$nnt_l, e$nnt_u)))
knitr::kable(res, caption = "Primary endpoint: effect of indomethacin vs placebo")
cat(sprintf("\nChi-square p-value: %.4f\n", chip))
```

```{r primary-plot, fig.height=4.4}
pd <- dat |> group_by(rx) |>
  summarise(rate = mean(pep_event), n = n(), ev = sum(pep_event), .groups = "drop")
ggplot(pd, aes(rx, rate, fill = rx)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%\n(%d/%d)", rate*100, ev, n)),
            vjust = -0.25, fontface = "bold", size = 4.2) +
  scale_fill_manual(values = PAL, guide = "none") +
  scale_y_continuous(labels = percent, limits = c(0, 0.235), expand = c(0, 0)) +
  annotate("segment", x = 1, xend = 2, y = 0.205, yend = 0.205, colour = INK) +
  annotate("text", x = 1.5, y = 0.222,
           label = sprintf("Absolute risk reduction %.1f points  |  NNT %.0f",
                           e$arr*100, e$nnt), fontface = "bold", colour = CORAL) +
  labs(title = "Pancreatitis after ERCP, by treatment arm",
       x = NULL, y = "Incidence of pancreatitis")
```

Indomethacin roughly **halved** the incidence of pancreatitis, from
`r sprintf("%.1f%%", e$p0*100)` on placebo to `r sprintf("%.1f%%", e$p1*100)`.
The absolute risk reduction of `r sprintf("%.1f", e$arr*100)` percentage points
translates into a **number needed to treat of about `r round(e$nnt)`**: treat
roughly `r round(e$nnt)` high-risk patients with one cheap suppository to prevent
one case of pancreatitis. That is an unusually efficient preventive effect.

# Adjusted analysis

Randomisation already balances the arms, but adjusting for baseline risk factors
gives a more precise estimate and shows the effect holds independent of them.

```{r adjusted}
fit <- glm(pep_event ~ rx + risk + age + gender + train,
           family = binomial, data = dat)
adj <- tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(label = recode(term,
    rxIndomethacin = "Indomethacin vs placebo", risk = "Risk score (per point)",
    age = "Age (per year)", genderMale = "Male vs female",
    trainYes = "Trainee involved"))
ggplot(adj, aes(estimate, fct_rev(label))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), colour = ACC, linewidth = 0.9) +
  geom_text(aes(label = sprintf("%.2f (%.2f to %.2f)", estimate, conf.low, conf.high)),
            vjust = -0.9, size = 3.2) +
  scale_x_log10(limits = c(0.2, 9), breaks = c(0.25, 0.5, 1, 2, 4)) +
  labs(title = "Adjusted odds ratios for pancreatitis after ERCP",
       subtitle = "Odds ratio below 1 means lower risk; horizontal bars are 95% CIs",
       x = "Adjusted odds ratio (log scale)", y = NULL)
```

```{r adjusted-or}
aor <- adj |> filter(term == "rxIndomethacin")
cat(sprintf("Adjusted OR for indomethacin: %.2f (95%% CI %.2f to %.2f), p = %.4f\n",
            aor$estimate, aor$conf.low, aor$conf.high, aor$p.value))
```

After adjusting for the composite risk score, age, sex and trainee involvement,
indomethacin remained associated with roughly **half the odds** of pancreatitis,
and the effect was statistically significant. The composite risk score was, as
expected, a strong independent predictor of the outcome.

# Subgroup analysis: who benefits?

A key clinical question is whether the benefit is consistent across patient
types, or concentrated in some. We estimate the relative risk within each
subgroup and test for an interaction (effect modification).

```{r subgroups, fig.height=5.6}
subs <- list(
  "Overall" = rep(TRUE, nrow(dat)),
  "Lower risk (<=2.5)"  = dat$risk_group == "Lower risk (<=2.5)",
  "Higher risk (>2.5)"  = dat$risk_group == "Higher risk (>2.5)",
  "SOD: no"  = dat$sod == "No",   "SOD: yes" = dat$sod == "Yes",
  "Prior pancreatitis: no"  = dat$prior_pep == "No",
  "Prior pancreatitis: yes" = dat$prior_pep == "Yes",
  "Female" = dat$gender == "Female", "Male" = dat$gender == "Male",
  "Trainee: no" = dat$train == "No", "Trainee: yes" = dat$train == "Yes")

rows <- lapply(names(subs), function(nm) {
  s <- dat[subs[[nm]], ]
  t <- table(factor(s$rx, levels = c("Placebo","Indomethacin")), s$pep)
  if (any(t[, "Yes"] == 0)) return(NULL)
  ee <- eff(t["Indomethacin","Yes"], t["Indomethacin","No"],
            t["Placebo","Yes"], t["Placebo","No"])
  data.frame(group = nm, rr = ee$rr, lo = ee$rr_l, hi = ee$rr_u,
             n = nrow(s))
})
sub_df <- do.call(rbind, rows)
sub_df$group <- factor(sub_df$group, levels = rev(sub_df$group))

ggplot(sub_df, aes(rr, group)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = lo, xmax = hi),
                  colour = ifelse(sub_df$group == "Overall", CORAL, ACC),
                  linewidth = 0.8) +
  geom_text(aes(label = sprintf("RR %.2f (n=%d)", rr, n)), hjust = -0.12,
            size = 3.1, nudge_y = 0.18) +
  scale_x_log10(limits = c(0.1, 5), breaks = c(0.25, 0.5, 1, 2)) +
  labs(title = "Relative risk of pancreatitis (indomethacin vs placebo) by subgroup",
       subtitle = "RR below 1 favours indomethacin; bars are 95% CIs",
       x = "Relative risk (log scale)", y = NULL)
```

```{r interaction}
p_risk  <- anova(glm(pep_event ~ rx * risk_group, binomial, dat), test = "LRT")$`Pr(>Chi)`[4]
p_sod   <- anova(glm(pep_event ~ rx * sod, binomial, dat), test = "LRT")$`Pr(>Chi)`[4]
p_prior <- anova(glm(pep_event ~ rx * prior_pep, binomial, dat), test = "LRT")$`Pr(>Chi)`[4]
cat(sprintf("Interaction p-values:  risk group = %.2f | SOD = %.2f | prior pancreatitis = %.2f\n",
            p_risk, p_sod, p_prior))
```

The point estimates favour indomethacin in essentially every subgroup, and no
interaction test is significant (all p > 0.05). In plain terms, the benefit is
**consistent across patient types**: there is no good evidence that any group
fails to benefit, so the treatment can be applied broadly to high-risk patients
rather than restricted to a subset.

# Results

```{r summary, echo=FALSE}
```

This section states the findings in the style of a clinical journal.

Among `r nrow(dat)` patients at high risk for post-ERCP pancreatitis randomised
to rectal indomethacin or placebo, the arms were well balanced at baseline
(Table 1). Post-ERCP pancreatitis occurred in `r a` of `r a+b`
(`r sprintf("%.1f%%", e$p1*100)`) patients given indomethacin, compared with
`r c` of `r c+d` (`r sprintf("%.1f%%", e$p0*100)`) given placebo (relative risk
`r sprintf("%.2f", e$rr)`; 95% CI, `r sprintf("%.2f", e$rr_l)` to
`r sprintf("%.2f", e$rr_u)`; P = `r sprintf("%.3f", chip)`). This corresponds to
an absolute risk reduction of `r sprintf("%.1f", e$arr*100)` percentage points
and a number needed to treat of `r round(e$nnt)` (95% CI, `r round(e$nnt_l)` to
`r round(e$nnt_u)`). The effect persisted after adjustment for the composite risk
score, age, sex and trainee involvement (adjusted odds ratio
`r sprintf("%.2f", aor$estimate)`; 95% CI, `r sprintf("%.2f", aor$conf.low)` to
`r sprintf("%.2f", aor$conf.high)`; P = `r sprintf("%.3f", aor$p.value)`). The
treatment effect was consistent across pre-specified subgroups, with no
significant interaction by risk group, sphincter of Oddi dysfunction or history
of prior pancreatitis.

# Conclusions and clinical implications

In this randomised, placebo-controlled trial, a single dose of rectal
indomethacin approximately halved the risk of post-ERCP pancreatitis in
high-risk patients, with a number needed to treat of about `r round(e$nnt)`. The
benefit was large, statistically robust, adjusted-analysis stable and uniform
across subgroups. Given that the intervention is inexpensive, single-dose and
widely available, these results support routine prophylactic rectal indomethacin
for patients at high risk of this complication.

# Limitations

The trial enrolled patients at elevated risk of pancreatitis, so the absolute
benefit and NNT apply to that population and would be smaller in average-risk
patients. The bleeding safety endpoint was too sparsely recorded in this dataset
to analyse, so this report addresses efficacy only. Subgroup analyses have
limited power and are exploratory. As with any single trial, confirmation in
other settings strengthens the case, which subsequent studies of this
intervention have provided.

---

*Built in R (dplyr, gtsummary, broom, ggplot2) from the `indo_rct` dataset in the
medicaldata package, a real published randomised controlled trial. Author:
Kingsley Amegah.*
